Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews

### 2025-04-21 analysis/paper/paper.qmd
gams

The question of archaeology's status as a science usually comes up in the context of what the discipline should or should not be. One of the first landmarks in tackling this question is the debate published by *Antiquity* between classical archaeologist Jacquetta Hawkes and palaeoanthropologist Glynn Isaac. @hawkesProperStudyMankind1968, advocating a humanistic archaeology, was concerned that scientific approaches to archaeology were causing researchers to be "swamped by a vast accumulation of insignificant, disparate facts, like a terrible tide of mud, quite beyond the capacity of any man to contain and mould into historical form". More optimistic about the integration of science and archaeology, @isaacWhitherArchaeology1971 counters that "New levels of precision in presenting data and in interpreting them can surely lead to briefer and more interesting technical reports as well as providing the basis for more lively literary portrayals of what happened in prehistory. Expanding on Isaac's perspective, @binford1962archaeology argued that archaeology should operate as a science after the model proposed by philosopher Carl Hempel, which prescribed hypothesis-driven approaches, leading to generalizable laws of human behavior. Counter-arguments came from numerous directions, notably @hodder1985postprocessual who rejected the quest for generalisations and instead argued that archaeology should be subjective and reflective, focussed on symbolic and relational meanings of material culture and the historical particularity of past human cultures. These debates, and the many more similar ones summarised by @martinon-torresArchaeologicalTheoriesArchaeological2013, have become a genre in archaeological writing that can be characterized as mostly based on personal observations, microscopic dissections of a handful of cherry-picked case studies of good or bad practice, and discussion of various philosophers and sociologists.

What has been missing from these debates is a macroscopic observation of what the majority of archaeologists are actually doing, and an empirical comparison to a broad spectrum of relatively harder and softer disciplines. At the 'hard' end of the spectrum (e.g. physics and chemistry), scholars more typically share a large set of established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results. At the 'soft' end of the spectrum (e.g. economics and psychology), the set of theories, facts, and methods on which there is widespread consensus is smaller, and agreement is slower and less frequently reached about the significance of new findings and the continuing relevance of previous work. Hardness and softness is a controversial distinction, in part because it is often used to imply a rank order of disciplines that encodes legitimacy, productivity, perceived value to society, and worthiness of funding [@DifferentAgenda2012; @coleHierarchySciences1983]. However, independent of these value judgments, empirical analysis of scholarly articles does indicate a spectrum of variation in practice linked to differing degrees of consensus in a discipline, for example in approaches to data visualisation [@clevelandGraphsScientificPublications1984; @smithScientificGraphsHierarchy2000]. Similarly, quantitative analysis of the frequency of positive results (ie. full or partial support for a research hypothesis) in publications is significantly correlated with hardness, consistent with a model where researchers in harder fields more readily accept any result their research produces, while those in softer fields have more freedom to choose which theories and hypotheses to test and how to interpret results [@fanelliPositiveResultsIncrease2010]. The hard-soft spectrum is also evident in surveys of how researchers view their own work relative to those in other fields [@biglanCharacteristicsSubjectMatter1973].

@@

over_time_long_colour_gams <- 
over_time_long_colour %>% 
  nest(.by = variable) %>% 
  mutate(mod_gam = lapply(data, 
                          function(df) gam(year ~ s(value, bs = "cr"), 
                                           data = df)))

over_time_long_colour_gams_summary <- 
over_time_long_colour %>% 
  nest(.by = variable) %>% 
  mutate(fit = map(data, ~mgcv::gam(year ~ s(value, bs = "cs"), data = .)),
         results = map(fit, glance),
         R.square = map_dbl(fit, ~ summary(.)$r.sq)) %>%
  unnest(results) %>%
  select(-data, -fit) %>% 
  select(variable, adj.r.squared)

@@

While these bibliometric variables provide several interesting insights into the status of archaeology as a science, via measurement of consensus, and are important for moving the debate beyond discussions of a small number of case studies, they miss a crucial factor that separates scientific practice from non-science. This is reproducibility, which, according to a report by the US National Science Foundation [@cacioppo2015social], "refers to the ability of a researcher to duplicate the results of a prior study using the same materials as were used by the original investigator. That is, a second researcher might use the same raw data to build the same analysis files and implement the same statistical analysis in an attempt to yield the same results... Reproducibility is a minimum necessary condition for a finding to be believable and informative." The importance of this factor can be traced to Irish chemist Robert Boyle (1627-1691), best known for his experiments with vacuum pumps [@shapin2011leviathan]. Boyle was concerned about the secrecy common among experimentalists in the 17th century and aimed to shift the culture from valuing direct in-person witnessing of scientific demonstrations towards meticulous written communications that were detailed enough to enable a reader to successfully undertake the experiment themselves, independent of the original author.

>>>>

The question of archaeology's status as a science usually comes up in the context of what the discipline should or should not be. One of the first landmarks in tackling this question is the debate published by *Antiquity* between classical archaeologist Jacquetta Hawkes and palaeoanthropologist Glynn Isaac. @hawkesProperStudyMankind1968, advocating a humanistic archaeology, was concerned that scientific approaches to archaeology were causing researchers to be "swamped by a vast accumulation of insignificant, disparate facts, like a terrible tide of mud, quite beyond the capacity of any man to contain and mould into historical form". More optimistic about the integration of science and archaeology, @isaacWhitherArchaeology1971 counters that "New levels of precision in presenting data and in interpreting them can surely lead to briefer and more interesting technical reports as well as providing the basis for more lively literary portrayals of what happened in prehistory. Expanding on Isaac's perspective, @binford1962archaeology argued that archaeology should operate as a science after the model proposed by philosopher Carl Hempel, which prescribed hypothesis-driven approaches, leading to generalizable laws of human behavior. Drawing on a different group of philosophers, @Smith_2017 argues for archaeology more specifically as a social science. Counter-arguments to archaeology as a science come from numerous directions, notably @hodder1985postprocessual who rejected the quest for generalisations and instead argued that archaeology should be subjective and reflective, focussed on symbolic and relational meanings of material culture and the historical particularity of past human cultures. These debates, and the many more similar ones summarised by @martinon-torresArchaeologicalTheoriesArchaeological2013, have become a genre in archaeological writing that can be characterized as mostly based on personal observations, microscopic dissections of a handful of cherry-picked case studies of good or bad practice, and discussion of various philosophers and sociologists.

What has been missing from these debates is a macroscopic observation of what the majority of archaeologists are actually doing, and an empirical comparison to a broad spectrum of relatively harder and softer disciplines. At the 'hard' end of the spectrum (e.g. physics and chemistry), scholars more typically share a large set of established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results. At the 'soft' end of the spectrum (e.g. economics and psychology), the set of theories, facts, and methods on which there is widespread consensus is smaller, and agreement is slower and less frequently reached about the significance of new findings and the continuing relevance of previous work. In sum, the hard-soft status is defined by the amount of consensus in a field, and the speed at which consensus is reached on new knowledge [@fanelliBibliometricEvidenceHierarchy2013]. Hardness and softness is a controversial distinction, in part because it is sometimes used to imply a rank order of disciplines that encodes legitimacy, productivity, perceived value to society, and worthiness of funding [@DifferentAgenda2012; @coleHierarchySciences1983]. On the other hand, analyzing the characteristics that lead to the hard-soft distinction can be useful for understanding the diversity of academic inquiry, such as how different fields approach knowledge and differences in what counts as evidence and modes of argument, where fruitful collaborations might be possible due to shared methods and assumptions, and for curriculum design to structure courses appropriately based on a field's typical ways of knowing.

Independent of these value judgments, empirical analysis of scholarly articles does support the hard-soft concept as a spectrum of variation in practice linked to differing degrees of consensus in a discipline, for example in approaches to data visualisation [@clevelandGraphsScientificPublications1984; @smithScientificGraphsHierarchy2000]. Similarly, quantitative analysis of the frequency of positive results (ie. full or partial support for a research hypothesis) in publications is significantly correlated with hardness, consistent with a model where researchers in harder fields more readily accept any result their research produces, while those in softer fields have more freedom to choose which theories and hypotheses to test and how to interpret results [@fanelliPositiveResultsIncrease2010]. The hard-soft spectrum is also evident in surveys of how researchers view their own work relative to those in other fields [@biglanCharacteristicsSubjectMatter1973].

@@

over_time_long_colour_gams <- 
over_time_long_colour %>% 
  nest(.by = variable) %>% 
  mutate(mod_gam = lapply(data, 
                          function(df) gam(value ~ s(year, bs = "cr"), 
                                           data = df)))

over_time_long_colour_gams_summary <- 
over_time_long_colour %>% 
  nest(.by = variable) %>% 
  mutate(fit = map(data, ~mgcv::gam(value ~ s(year, bs = "cs"), data = .)),
         results = map(fit, glance),
         R.square = map_dbl(fit, ~ summary(.)$r.sq)) %>%
  unnest(results) %>%
  select(-data, -fit) %>% 
  select(variable, adj.r.squared)

@@

While these bibliometric variables provide several interesting insights into the status of archaeology as a science, via measurement of consensus, and are important for moving the debate beyond discussions of a small number of case studies, they miss a crucial factor that separates scientific practice from non-science. This is reproducibility, which, according to a report by the US National Science Foundation [@cacioppo2015social], "refers to the ability of a researcher to duplicate the results of a prior study using the same materials as were used by the original investigator. That is, a second researcher might use the same raw data to build the same analysis files and implement the same statistical analysis in an attempt to yield the same results... Reproducibility is a minimum necessary condition for a finding to be believable and informative." Scientific reproducibility is linked to consensus because if more researchers provide sufficient detail for others to reproduce their results, then consensus on new knowledge can more often and more rapidly established. The importance of this factor can be traced to Irish chemist Robert Boyle (1627-1691), best known for his experiments with vacuum pumps [@shapin2011leviathan]. Boyle was concerned about the secrecy common among experimentalists in the 17th century and aimed to shift the culture from valuing direct in-person witnessing of scientific demonstrations towards meticulous written communications that were detailed enough to enable a reader to successfully undertake the experiment themselves, independent of the original author.
